Performance of Individual Audit Filters in Predicting Opportunities for Improvement

### 2024-06-08 manuscript.Rmd
update

Introduction
============
Trauma, defined as a physical injury and
the body's associated response, accounts for 8-9%
of global deaths annually [@world2014injuries; @roth2018global], mainly affecting younger individuals in low- and middle-income countries attributed road traffic injuries [@vos2020global;
@roth2018global; @SweTrau2022Arsrapport]. Furthermore, trauma is responsible for the top cause for disability-adjusted life-years
(DALY) [@vos2020global] and associated with 36% not returning to work within one year
[@david2022measuring] and a higher
propensity towards depression, post-traumatic stress disorder (PTSD) and
negative effects on cognitive function [@kessler1995posttraumatic;
@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts; @visser2017course]. Trauma is
associated with . Attempts have been made to estimate the global
cost of trauma, which is a complicated task due to its variability.
However, it is estimated that the cost of road traffic injuries alone
accounts for 2% of gross domestic product (GDP) in high-income countries
[@world2014injuries].

## Trauma Care Systems

In order to reduce impact trauma care systems are in place. 
Trauma care is a multidisciplinary, complex and time-critical health
care service provided by hospitals
[@American2022resources]. Trauma care is divided into three components:
pre-hospital, in-hospital and post-hospital. The pre-hospital component
constitutes first responders, basic life support and advanced life
support [@world2005prehospital]. The in-hospital component is defined as
the moment a trauma patient is admitted to a health care facility, often
a trauma centre. The post-hospital component involves the rehabilitation
and recovery of treated trauma patients.

Trauma centres have a significantly lower mortality rate compared to
non-trauma centres, and the maturity of a trauma centre correlates with
lower mortality, demonstrating the importance of high-quality,
specialised trauma care [@alharbi2021effectiveness; @mann1999systematic;
@candefjord2020mortality]. Despite this, even established trauma care
systems are prone to errors due to the complex nature of trauma and its
susceptibility to delays [@nikouline2021errors; @vioque2014classifying;
@o2013opportunities]. Studies between 1990 and 2014 have shown
a pooled preventable death rate of 20%, with more recent studies showing
a lower preventable death rate [@kwon2014pooled]. For example, a recent
study in a Swedish Level I trauma centre found a preventable death rate
of 4% [@ghorbani2018analysis].

## Quality Improvement Programmes

To minimise errors in trauma care, quality improvement (QI) programs
have been established by the World Health Organization (WHO) and the
International Association for Trauma Surgery and Intensive Care (IATSIC)
[@world2009guidelines]. These programs form the core of current trauma
care systems and their development around the world
[@American2022resources]. QI programs consist of a series of recommended techniques aiming to
facilitate discussion and analysis of the trauma process in order to
take corrective action for future success. The following techniques are
commonly used: a multidisciplinary mortality and morbidity conference
(MMC), a preventable death review panel, and the use of AFs also known
as quality indicators [@world2009guidelines].

@@

@@

Discussion
==========

@@

<!-- Give a short conclusion, corresponding to the aim -->

@@

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

>>>>

Introduction
============
Trauma, defined as a physical injury and
the body's associated response, accounts for 8-9%
of global deaths annually [@world2014injuries; @roth2018global], mainly affecting younger individuals in low- and middle-income countries attributed to mainly road traffic injuries [@vos2020global;
@roth2018global; @SweTrau2022Arsrapport]. Furthermore, trauma is responsible for the top cause for disability-adjusted life-years (DALY) [@vos2020global] and a higher propensity towards mental disorders and cognitive decline [@kessler1995posttraumatic;
@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts; @visser2017course]. This negatively affects individuals, communities and society as a whole [@newnam2014impacts].It is estimated that the cost of road traffic injuries alone
accounts for 2% of gross domestic product (GDP) in high-income countries
[@world2014injuries].

To reduce trauma burden, quality improvement (QI) programs
have been established by the World Health Organization (WHO) and the
International Association for Trauma Surgery and Intensive Care (IATSIC)
[@world2009guidelines]. These programs form the core of current trauma
care systems and their development around the world
[@American2022resources]. QI programs consist of a series of recommended techniques aiming to
facilitate discussion and analysis of the trauma process in order to
take corrective action for future success. The following techniques are
commonly used: a multidisciplinary mortality and morbidity conference
(MMC), a preventable death review panel, and the use of AFs also known
as quality indicators [@world2009guidelines].

Trauma care is a multidisciplinary, complex and time-critical health
care service provided by hospitals, often trauma centres.
[@American2022resources].
Trauma centres have a significantly lower mortality rate compared to
non-trauma centres, and the maturity of a trauma centre correlates with
lower mortality, demonstrating the importance of high-quality,
specialised trauma care [@alharbi2021effectiveness; @mann1999systematic;
@candefjord2020mortality]. Despite this, even established trauma care
systems are prone to errors due to the complex nature of trauma and its
susceptibility to delays [@nikouline2021errors; @vioque2014classifying;
@o2013opportunities]. Studies between 1990 and 2014 have shown
a pooled preventable death rate of 20%, with more recent studies showing
a lower preventable death rate [@kwon2014pooled]. For example, a recent
study in a Swedish Level I trauma centre found a preventable death rate
of 4% [@ghorbani2018analysis].

@@

##Participants
After excluding patients
under 15 years old and patients not screened for OFIs, a total of 8,309
individuals were included in the study. 7,797 out of 8,309 (93.8%)
patient cases did not have an OFI and 512 out of 8,309 (6.2%) had an
OFI. Figure 1 presents a flowchart showcasing the exclusion and inclusion of
the patient cases from the trauma registry.

Most trauma patients were men (n = 5,755, 69%). Patient characteristics
varied depending on whether patients had an OFI or not. Respectively,
the mean age was significantly higher\
`r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` vs
`r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)`,
`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")`
and
the ISS was significantly higher
`r inline_text(tableOne, variable = ISS, column = stat_2)` vs
`r inline_text(tableOne, variable = ISS, column = stat_1)`,
`r inline_text(tableOne, variable = ISS, column = "p.value")`. The
number of intubations in the emergency department was significantly
higher
`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_2)`
vs
`r inline_text(tableOne, variable = ed_intubated, level = "No", column = stat_1)`,
`r inline_text(tableOne, variable = ed_intubated, level = "No", column = "p.value")`.
Patients with OFI had longer times to definitive treatment from hospital
arrival compared to patients without OFI
`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)`
vs
`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1)`,
`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = "p.value")`.
The time to the first CT was also longer in patients with OFIs compared
to patients without
`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2)` vs
`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1)`,
`r inline_text(tableOne, variable = dt_ed_first_ct, column = "p.value")`.

The number of OFIs were highest in patients admitted to the intensive
care unit
`r inline_text(tableOne, variable = host_care_level, level = "Intensive care unit", column = stat_2)`
and surgical ward
`r inline_text(tableOne, variable = host_care_level, level = "Surgical ward", column = stat_2)`.

Gender `r inline_text(tableOne, variable = Gender, column = "p.value")`,
the mean systolic blood pressure
`r inline_text(tableOne, variable = ed_sbp_value, column = "p.value")`,
death at 30 days
`r inline_text(tableOne, variable = res_survival, column = "p.value")`
and trauma team activation
`r inline_text(tableOne, variable = Tr_Nivå, column = "p.value")` was not
significantly correlated with OFI.

Notably, there was a statistical significant correlation between GCS at
ED and OFI (p \< 0.001).

Table 2 shows the missing data for each AF. As mentioned in the method
section, these values were interpreted as "not flagged" with an AF. The
variables with the highest amount of missing data were "time to
definitive treatment" (n = 5,990, 72.1%) and "time to first CT" (n =
1,012, 12.2%). The smallest amount of missing data was found in
"SBP\<90" (n = 13, 0.2%).

## Performance of Individual Audit Filters

The performance of each AF was determined through a calculation of
sensitivity and specificity which is demonstrated in Table 4.

The number of patient cases flagged by each individual AF varied between
107 for "GCS\<9 and not intubated" and 3907 for "\>30 min until CT". The
AF with the highest sensitivity was "\>30 min until CT" (59.2%, 95% CI
[55.2-63.8]). The lowest sensitivity was seen in "GCS\<9 and not
intubated" (1.2%, 95% CI [0.2-2]). The AF with the highest specificity
was "GCS\<9 and not intubated" (98.7%, 95% CI [98.5-98.9]). The lowest
specificity was seen in "\>30 min until CT" (53.8%, 95% CI [52.7-54.8]).
Notably, the confidence intervals varied between sensitivity and
specificity, with sensitivity displaying a greater interval.

Table 5 summarises the performance of each individual AF in identifying
OFIs in trauma care by AUC. Four of the 11 AFs, "SBP\<90", "dead at 30
days", "GCS\<9 and not intubated" and "CPR and thoracotomy" showed a
performance in predicting OFIs equal to that of coincidence (AUC = 0.5).
The highest AUC was seen in "\>60 min until first intervention" (0.63,
95% CI [0.61-0.65]). All AFs showed significantly lower AUC compared to
the minimum standard for diagnostic tests (AUC = 0.8) (p \< 0.001).

Table 2: Comparison of missing values for each AF. Missing values is
shown as a number and percentage of the total number of patient cases
included in the study (n = 8309).\
`r tableTwo` Table 3: Patient characteristics of individuals with and
without an OFI. Missing values are shown only when they exist.
`r tableOne` Table 4: Comparison of performance for each AF in
predicting OFIs. Performance is defined by sensitivity (%) and
specificity (%). A 95% confidence interval is shown together with each
performance indicator. The number of cases flagged by each filter is
displayed.\
`r tableThree` Table 5: Comparison of performance for each AF in
predicting OFIs. Performance is defined by the area under the receiver
operating characteristics curve. P-values show the significance of AUC
compared to coincidence (AUC = 0.5) and to the minimum standard for
diagnostic tests (AUC = 0.8) [@nahm2022receiver]. The p-values were
calculated using an empirical bootstrap test. Furthermore, a 95%
confidence interval was used. `r tableFour`

```{r fig1, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE, error=FALSE}
knitr::include_graphics("ofi-flowchart.pdf")
```

Figure 1: Flowchart of the patient selection process. Shows the patients
excluded and included in this study.
<!--

@@

Discussion
==========
This study showed that AFs do not perform well in predicting OFIs. No single AF presented with a performance of value in predicting OFI in trauma care patients. Some AF showed a correlation with OFI but not enough to be considered useful.

Patient characteristics are similar to previous studier showing twofold prevalence of male trauma patients compared to females [@roth2018global;@SweTrau2022Arsrapport]. Previous studies have shown

A previous study conducted at Karolinska University Hospital by Albaaj
et al. [@albaaj2023patient] sought to identify patient factors
associated with OFIs and found that a higher ISS, longer times to CT,
longer times to first intervention and higher levels of care was
correlated with OFIs. These findings correspond with our analysis of
patient characteristics and AFs linked to these factors.

Higher ISS and higher levels of care in patients with OFIs as opposed to
patients without OFIs could be explained by the greater complexity of
such trauma cases. Greater complexity requires more interventions which
could allow for more mistakes [@albaaj2023patient].

Notably, GCS at ED showed a correlation with OFI even though the median
GCS and its IQR was identical for individuals with OFI and without. Upon
further analysis this could be explained by the difference in
distribution outside the IQR.

AFs generally exhibited a high specificity relative to their sensitivity
which may be attributed to the outcome being rather uncommon, whilst
every individual AF misses the majority of OFIs. This could suggest that
the AFs are too rigid in their criteria. On the contrary, a too
inclusive criteria might flag several patients, who do not necessarily
have an OFI [@willis2008interpreting].

In 2008, Willis et al. [@willis2008interpreting] assessed 14 audit
filters, finding only three to be associated with an increased risk of
poor outcomes. One AF, "GCS\<9 and not intubated", also used in this
study, similarly showed no significant ability to predict OFIs. An
explanation for this could be that an abnormal GCS as a result from
intoxication is not comparable to an abnormal GCS due to head injury for
example, suggesting that the AF might be more reflective of clinical
decision rather than quality of care [@willis2008interpreting]. However,
according to Sanddal et al. [@sanddal2011analysis] airway management is
associated with OFI suggesting that an alternative audit filter related
to airway management might potentially perform differently.

Both AFs measuring delay: "\>60 until first intervention" and "\>30min
until first CT", showed a correlation with OFI. This is comparable to
what Teixeira et al. [@teixeira2007preventable] demonstrated, with delay
correlating to preventable death. Furthermore, Ghorbani et al.
[@ghorbani2018analysis] identified delay in treatment, delay to CT, and
clinical judgement errors as common errors in trauma care. A more recent
study in Japan by Yamamoto et. al [@yamamoto2023immediate] showed a
significant reduction in mortality when patients received a whole-body
CT within 10 minutes of arrival at the ED. It is important to note that
OFI in this study was not only a reflection of mortality but a broader
concept encompassing different aspects of care quality. If anything, it
highlights the difference between hospitals, which AFs are in use, and
how they are being assessed.

AFs that present with a high risk of inevitable death showed poor"Dead
at 30 days" and "CPR and thoracotomy", could be explained by inevitable
mortality, thus presenting with a smaller time frame for mistakes to be
made.

##Interpretation

AFs are static in their criteria, contrary to health care in general.
Consequently, the utility of AFs may diminish if confronted with
emerging challenges or upon resolution of the quality gaps they were
designed for. The static nature of AFs highlight their need for
consistent and evidence-based review to determine current usefulness
[@santana2014development]. However, it is worth pointing out that the
review process might present with costs greater than the potential
benefit from AFs.

AFs face noteworthy challenges due to over-reliance, bias and
confounders. Firstly, while AFs may help in determining the existence of
OFI, they could overshadow other factors that may contribute to that
analysis, leading to over-reliance. Secondly, devising AFs comes with an
unintended bias. They might disproportionally affect certain groups of
people failing to recognise patient cases that fall outside the scope of
AF criteria [@willis2008interpreting]. For example, AFs primarily focus
on aspects of trauma care mostly relevant to medical doctors despite the
interdisciplinary nature of trauma care. This is especially
counterintuitive given the broad definition of OFIs. Lastly, confounding
variables adds to the uncertainty of AFs performance. For example,
patients receiving the same interventions at different time points may
do so as a function of injury severity not quality of care
[@willis2008interpreting].

## Limitations

Opportunity for improvement, while dichotomously defined, includes a
diverse set of outcomes. This convolutes the process of creating AFs
that perform well in predicting OFIs. The current screening system for
OFIs might also introduce bias since the AFs would favour the
identification of some, but not all, errors.

Each AF was individually related to OFI. Consequently, even if an AF
demonstrated a true positive correlation with an OFI, instances where
the same patient case had multiple AFs introduce a degree of uncertainty
regarding which specific AF accurately predicted the outcome.

The number of missing values varied greatly among the AFs. As stated
above, missing values from each AF were coded as "not flagged" by an AF.
This means that some instances where patient cases were not flagged,
might actually meet the AF criteria, but for some reason not documented.
This reduces the validity of some AFs that featured high amounts of
missing data.

## Generalizability

AFs are widely used but generalizability is complicated due to the lack
of common grounds on what they are, what purpose they serve and how they
shall be used [@stelfox2010quality; @stelfox2011evidence;
@berg2022perceived].

Our findings suggest that AFs perform poorly in predicting OFIs. The AFs
with performance exceeding coincidence (AUC = 0.5) might be a beneficial
addition to an already existing quality improvement programme or could
be used to fine-tune AFs already in use to plug hospitals' quality gaps.
Due to the study being conducted in a single, trauma level I centre in
Stockholm it is unrealistic to assume applicability of these findings at
other hospitals, in other countries, facing unique challenges
[@world2009guidelines]. For example, Low and middle income countries
seem to benefit from certain AFs that high income countries do not and
vice versa [@berg2022perceived].

## Strengths

As stated in the method section, some AFs were manually created from
registry data instead of relying on the selection process by the
specialised nurse (see table 1). This, in order to minimise
misclassification bias.

Furthermore, there has been very limited high quality evidence
supporting the efficacy of AFs in improving trauma care
[@evans2009audit; @stelfox2010quality] and predicting OFI
[@sanddal2011analysis; @roy2017learning]. This study shows comparable
results. Previous studies also focus primarily on AFs' ability to reduce
mortality [@chadbunchachai2001comparative; @chadbunchachai2003study]. In
this study the morbidity of patient cases is also taken into account
using OFIs as an umbrella term that includes factors ranging from bad
documentation to preventable death.

From 2017 all trauma patients admitted to Karolinska University Hospital
were included in the registry reducing the risk of bias in inclusion of
patient cases.

## Future Studies

With the ongoing academic, political, and public attention towards
patient safety and quality of healthcare [@evans2009audit], it is
important to discover optimal methods for monitoring and enhancing the
care provided within trauma systems.

Given the expanding field of artificial intelligence and machine
learning, newer studies should focus on using these tools to screen for
patient cases at risk of OFIs. This will diminish misclassification
rates and ensure consistent and sustained high performance, attributable
to an algorithm's capability to adapt and learn from the prevailing
context of trauma care. This approach is contemporary but not
unexplored. A preprint is currently available by Attergrim et al.
[@attergrim2023predicting] exploring the idea of machine learning models
as a screening tool for potential OFIs. They found that machine learning
models outperformed the AFs currently used at Karolinska University
Hospital in predicting OFIs.

Should the use of AFs remain, it is important to ensure its performance
by routinely updating them in order to better reflect the current
quality gaps.
<!--

@@

AFs are widely used as a tool in quality improvement programmes and
should reflect the current gap in trauma care quality. Strong evidence
supporting their efficacy in quality improvement is scarce. In this
study, no audit filter seemed to show promising performance in
predicting OFIs with AUC ranging from 0.5 to 0.63. These results
highlight the importance of continually evaluating the relevance of AFs.
However, this process takes time and resources and using an alternative
to AFs is recommended.

@@
